Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-07-25 tex/scam/optimization.tex
edits on qp formulation

Let $\bds{x}_{i}\in\mathbb{R}^{p}$ be the covariate, let $y_{i}$ be the
response and let $\epsilon_{i}$ be the mean zero noise. The regression function $f(\cdot)$ we estimate is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension and a scalar offset $\mu$.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
an univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      f_{i'k} \geq f_{ik} + \beta_{ik}(x_{i'k}-x_{ik}) \quad
      \textrm{for all $i,i' = 1,\ldots, n$,}
\end{equation}
where $f_{ik}\coloneqq f_{k}(x_{ik})$ and $\beta_{ik}$ is the
subgradient at point $x_{ik}$. We apparently need $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints.
In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to the  optimization
\begin{equation}
\begin{split}
       \min_{f,\beta,\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( y_{i}-\mu - \sum_{k=1}^{p}f_{ik}\Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|f_k\|_{\infty} \\
       \textrm{subject to} &\;\; \textrm{for all $k=1,\ldots, p$:}\\
       & \;\; f_{(i+1)_k k} = f_{(i)_k k} +
       \beta_{(i)_k k}(x_{(i+1)_k k}-x_{(i)_k k}),\;\textrm{for $i=1,\ldots, n-1$}\\
       & \;\; \sum_{i=1}^{n}f_{ik}=0,\\
       & \;\; \beta_{(i+1)_k k} \geq \beta_{(i)_k k}\;\textrm{for $i=1,\ldots, n-1$}.
\end{split}
\label{np}
\end{equation}
Here $f_k$ denotes the vector $f_k = (f_{1k}, f_{2_k}\ldots, f_{nk})^T\in\reals^n$
and $\{(1)_k,(2)_k,\ldots,(n)_k\}$ is a reordering of $\{1,2,\ldots,n\}$
such that 
\begin{equation}
x_{(1)_k k} \leq{} x_{(2)_k k} \leq \cdots \leq{} x_{(n)_k k}.
\end{equation}

@@

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
would be computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the term
$(Y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P \uds{f}_k = \diag(P \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zero except $-1$ at in position $(i)_k$ and $1$ in
position $(i+1)_k$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zero except for a $1$ 
at position $i$ and $-1$ at position $i+1$.  We denote by
$\diag( v )$ the diagonal matrix whose diagonal entries are $v$.

% \begin{equation}
% \label{eqn:opt_1d}
% \begin{split}
%        \min_{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot},\gamma_{k}} &
%              \ \frac{1}{2n}\sum_{i=1}^{n}\Bigl((Y_{i}-\bar{Y}
%                 -\sum_{r\neq{k}}f_{ri})-f_{ki}\Bigr)^{2} 
%                       + \lambda\gamma_{k} \\
%         \textrm{such that} & \ f_{k(i+1)} = f_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}),\\
%         &\ \beta_{k(i+1)} \geq \beta_{k(i)}, \ -\gamma_{k}\leq f_{ki}\leq\gamma_{k}\\
%         &\  \sum_{i=1}^{n}f_{ki}=0, \ (\forall i).
% \end{split}
% \end{equation}
The extra variable $\gamma_{k}$ is introduced to deal with the $\ell_{\infty}$ norm. This QP subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse structure, 
which can be solved efficiently using optimization packages (e.g., MOSEK: \verb+http://www.mosek.com/+).  We cycle through all feature dimensions ($k$) from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few cycles. We also implemented an ADMM solver for (\ref{np}), but found that it is not as efficient as this QP solver.

After optimization, the function estimator for any input data $\bds{x}_j$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
\nonumber
      f(\bds{x}_j) & = \sum_{k=1}^{p}f_k(x_{kj})+\mu 
= \sum_{k=1}^{p}\max_{i} \{f_{ki}+\beta_{ki}(x_{kj}-x_{ki})\} +
      \mu.
\end{split}
\end{equation}

The univariate concave function estimation is a straightforward modification of optimization~\ref{opt:1d_compact}. We need only modify one linear inequality constraint to enforce that the subgradients must be non-increasing: $\beta_{k(i+1)} \leq \beta_{k(i)}$.

>>>>

Let $\bds{x}_{i}\in\mathbb{R}^{p}$ be the covariate, let $y_{i}$ be the
response and let $\epsilon_{i}$ be the mean zero noise. The regression function $f(\cdot)$ we estimate is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension and a scalar offset $\mu$.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
an univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      f_{i'k} \geq f_{ik} + \beta_{ik}(x_{i'k}-x_{ik}) \quad
      \textrm{for all $i,i' = 1,\ldots, n$,}
\end{equation}
where $f_{ik}\coloneqq f_{k}(x_{ik})$ and $\beta_{ik}$ is the
subgradient at point $x_{ik}$. This apparently requires $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints.
In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to the  optimization
\begin{equation}
\begin{split}
       \min_{f,\beta,\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( y_{i}-\mu - \sum_{k=1}^{p}f_{ik}\Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|f_k\|_{\infty} \\
       \textrm{subject to} &\;\; \textrm{for all $k=1,\ldots, p$:}\\
       & \;\; f_{(i+1)_k k} = f_{(i)_k k} +
       \beta_{(i)_k k}(x_{(i+1)_k k}-x_{(i)_k k}),\;\textrm{for $i=1,\ldots, n-1$}\\
       & \;\; \sum_{i=1}^{n}f_{ik}=0,\\
       & \;\; \beta_{(i+1)_k k} \geq \beta_{(i)_k k}\;\textrm{for $i=1,\ldots, n-1$}.
\end{split}
\label{np}
\end{equation}
Here $f_k$ denotes the vector $f_k = (f_{1k}, f_{2_k}\ldots, f_{nk})^T\in\reals^n$
and $\{(1)_k,(2)_k,\ldots,(n)_k\}$ are the indices in the sorted ordering
of the values of coordinate $k$:
\begin{equation}
x_{(1)_k k} \leq{} x_{(2)_k k} \leq \cdots \leq{} x_{(n)_k k}.
\end{equation}

@@

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
is computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the squared error term
$(y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P \uds{f}_k = \diag(P \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zeros except for a $-1$ in position $(i)_k$ and a $1$ in
position $(i+1)_k$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zeros except for a $1$ 
in position $i$ and a $-1$ in position $i+1$.  We denote by
$\diag( v )$ the diagonal matrix with diagonal entries $v$.
The extra variable $\gamma_{k}$ is introduced to impose the
regularization penalty involving the $\ell_{\infty}$ norm.

% \begin{equation}
% \label{eqn:opt_1d}
% \begin{split}
%        \min_{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot},\gamma_{k}} &
%              \ \frac{1}{2n}\sum_{i=1}^{n}\Bigl((Y_{i}-\bar{Y}
%                 -\sum_{r\neq{k}}f_{ri})-f_{ki}\Bigr)^{2} 
%                       + \lambda\gamma_{k} \\
%         \textrm{such that} & \ f_{k(i+1)} = f_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}),\\
%         &\ \beta_{k(i+1)} \geq \beta_{k(i)}, \ -\gamma_{k}\leq f_{ki}\leq\gamma_{k}\\
%         &\  \sum_{i=1}^{n}f_{ki}=0, \ (\forall i).
% \end{split}
% \end{equation}

This QP
subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse
structure, which can be solved efficiently using optimization
packages. In our experiments we use {\sc mosek} (\href{http://www.mosek.com/}{www.mosek.com}).  We cycle through
all covariates $k$ from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few
cycles. We also implemented an ADMM solver for (\ref{np})
\citep{Boyd:admm}, but found
that it is not as efficient as this QP solver.

After optimization, the function estimates for an input vector $\bds{x}$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
      \hat f(\bds{x}) & = \sum_{k=1}^{p} \hat f_k(x_{j})+ \hat \mu 
= \sum_{k=1}^{p}\max_{i} \bigl\{\hat f_{ik}+ \hat \beta_{ik}(x_{k}-x_{ik})\bigr\} +
      \hat \mu.
\end{split}
\end{equation}

The univariate concave function estimation is a straightforward
modification of optimization~\ref{opt:1d_compact}. It is only
necessary to modify the linear inequality constraints so that the subgradients are
non-increasing: $\beta_{(i+1)_kk} \leq \beta_{(i)_kk}$.
